Refunctionalization of Abstracting Abstract Machines (Functional Pearl)

### 2018-03-17 paper.tex
2.3.1

  In this paper, we show their correspondence and how to syntactically transform small-step
  abstract abstract machines into big-step abstract definitional
  interpreters.
  The transformations include linearization, fusing, disentangling, refunctionalizing,
  and monadification (or un-CPS to direct-style). Linearization expresses
  non-deterministic choices by first-order data types, after which refunctionalization
  sequentializes the evaluation order by higher-order functions.
  All transformations properly handle the collecting semantics and the
  nondeterminism of abstract interpretation.
  %After each transformation, we also obtain an intermediate form of abstract interpreter.

@@

Abstracting abstract machines (AAM) is a systematic methodology proposed by \citeauthor{van2012systematic}
for constructing abstract interpreters of higher-order functional languages \cite{van2012systematic, van2010abstracting}. 
Given a concrete small-step abstract machine, (e.g, the CESK machine, Krivine's machine or etc), 
by allocating continuations in the store and bound both the value addresses and continuation
addresses to be finite, we obtain an abstract interpreter with finite state space which 
can be used for performing sound static analysis.
One can further instantiate different polyvariant control-flow analysis by using different address allocators \cite{Gilray:2016:ACP:2951913.2951936}.

Applying the same idea on big-step definitional interpreters, \citeauthor{darais2017abstracting} 
build abstract definitional interpreters (ADI) that are written in monadic style \cite{darais2017abstracting}
By changing the underlying
monads, the definition of interpreter has no change, but we can recover different semantics 
including the concrete semantics and various abstract semantics.

Broadly speaking, abstract abstract machines and abstract definitional interpreters are both
abstract interpreters. They are obtained by applying a combination of abstractions to their
concrete counterparts, abstract machines and definitional interperters, perspectively.
The relations between abstract machines, definitional interpreters, and monadic interpreters 
is intensively studied by Danvy and his collaborator in the past years 
\cite{Ager:2003:FCE:888251.888254, Danvy:2001:DW:773184.773202, danvy2004refocusing, Danvy:2008:DIP:1411204.1411206, AGER2004223, ager2005functional, Danvy:2006:RW:2171265.2171268}.

But in terms of their abstract semantic artifacts, 
the relations between small-step abstract abstract machines and big-step abstract 
definitional interpreters, and the question that how can we interdrive each other is not
well studied.
In this functional peral, we practical answer this question by presenting a series of 
syntactical transformations on the program from small-step abstract abstract machines 
to big-step abstract definitional interpreters.

\subsection{outline}

@@

\begin{itemize}
\item We show how \textbf{linearization} (Section~\ref{linear}) can be used to...
\item In Section~\ref{fusing}, we then present the \textbf{fusing} transformation, and present...
\item Section~\ref{disen} discusses \textbf{disentangling}, which shows...
\item \textbf{Refunctionalization} is shown in Section~\ref{refunc}, and brings a new understanding 
  of pushdown control-flow analysis. In this section, we present an abstract interpreter obtained 
  from refunctionalizing a small-step abstracting abstract machine which naturally exhibits the 
  pushdown control-flow property (as refunctionalization requires abstract execution to be sequentialized).
\item The last transformation we show is that of transforming a definitional abstract interpreter 
  to a \textbf{direct-style} interpreter (Section~\ref{directstyle}).

@@

These transformations are used throughout the paper, with refunctionalization and
defunctionalization of abstract interpreters playing important roles for the call stack of
the analyzed language. All of this insight allows the revelation of
precisely \emph{why} small-step AAM does not have the pushdown control-flow property
(without extra effort), as well as explains why ADI naturally extends pushdown control-flow
property from its defining language.

>>>>

  In this paper, we show their correspondence and how to syntactically transform small-step
  abstract abstract machines into big-step abstract definitional
  interpreters.
  The transformations include linearization, fusing, disentangling, refunctionalizing,
  and monadification (or un-CPS to direct-style). Linearization expresses
  nondeterministic choices into first-order data types, after which refunctionalization
  sequentializes the evaluation order by higher-order functions.
  All transformations properly handle the collecting semantics and the
  nondeterminism of abstract interpretation.
  %After each transformation, we also obtain an intermediate form of abstract interpreter.

@@

Abstracting abstract machines (AAMs) and abstract definitional interpreters (ADIs) are, broadly speaking, both examples of abstract interpreters. \note{JD: need more bullet points here and I can fill this in}

@@

\begin{itemize}
\item We show how \textbf{linearization} (Section~\ref{linear}) can be used to...
\item In Section~\ref{fusing}, we then present the \textbf{fusing} transformation, and present...
\item Section~\ref{disen} discusses \textbf{disentangling}, which shows...
\item \textbf{Refunctionalization} is shown in Section~\ref{refunc}, and brings a new understanding of pushdown control flow analysis. In this section, we present an abstract interpreter obtained from refunctionalizing a small-step abstracting abstract machine which naturally exhibits the pushdown control-flow property (as refunctionalization requires abstract execution to be sequentialized).
\item The last transformation we show is that of transforming a definitional abstract interpreter to a \textbf{direct-style} interpreter (Section~\ref{directstyle}).

@@

These transformations are used throughout the paper, with refunctionalization and
defunctionalization of abstract interpreters playing important roles for the call stack of
the analyzed language (i.e., Scala)\note{right?}. All of this insight allows the revelation of
precisely \emph{why} small-step AAM does not have the pushdown control-flow property
(without extra effort), as well as explains why ADI naturally extends pushdown control-flow
property from its defining language.
